---
title: "Getting Our Feet Wet With The forcats Package"
date: "Last Updated `r Sys.Date()`"
fontsize: 16pt
output:
  html_document:
    theme: united
    df_print: paged
---
 <br> 

This is the third and final `r emo::ji("sleepy")` part of the hands-on practice for today. If it is any consolation, this one is pretty brief because the *forcats* `r emo::ji("cat")` `r emo::ji("cat")` `r emo::ji("cat")` `r emo::ji("cat")` is kind of a one-trick pony... (pause for laughter... and applause?)

 * [Strings](https://r4ds.had.co.nz/strings.html)
 * [Factors](https://r4ds.had.co.nz/factors.html)
 * [Dates and Times](https://r4ds.had.co.nz/dates-and-times.html)

At any rate, you read about the *forcats* package today so let's practice a little to begin getting a feel for some of the more commonly used functions therein. We are going to load the dataset used in the *forcats* reading for today, which is just an excerpt from the [General Social Survey](https://gss.norc.org/About-The-GSS). Because these data are stored as a tibble object, we can see that many of them are factors^[Categorical data that can take on only one of a finite set of values, which are called levels.]. Let's use the `fct_count` function to see how many of the 21,438 records fall into various categories 

<br>

```{r label="Load R Packages", include= FALSE}

options(repos=c(CRAN="https://mirrors.nics.utk.edu/cran/")) 

library(tidyverse)
library(lubridate)
library(emo)
library(janitor)
library(jsonlite)

```



```{r label="Summary Of Respondents By Religious Affiliation"}

data(gss_cat)

fct_count(gss_cat$relig, sort = TRUE, prop = TRUE)

```

<br> 

As you probably deduced, the **sort argument** returns the results in descending order and the **prop argument** gives us the proportion of the total in each category. We haven't introduced the *ggplot2* package yet (next week `r emo::ji("grinning")`), but we can use the barplot code from prior examples to visualize these data. 

```{r label="Barchart Of Religious Affiliation"}

ggplot(data = gss_cat) + 
  geom_bar(stat = "count", aes(x = as_factor(relig)), fill = "orange") + 
  labs(x="", y = "Number of Respondents") + 
  theme(axis.text.x = element_text(size = 8, angle = 90))

```

There are an awful lot of categories here with not many respondents. Let's combine some of the categories and recreate the barplot... `r emo::ji("silly")`

```{r label="Better Barchart Of Religious Affiliation"}

gss_cat <- gss_cat %>%
  mutate(small.relig = fct_recode(relig,
                                  "Christian" = "Protestant", 
                                  "Christian" = "Orthodox-christian", 
                                  "Christian" = "Catholic"))

ggplot(data = gss_cat) + 
  geom_bar(stat = "count", aes(x = as_factor(small.relig)), fill = "blue") + 
  labs(x="", y = "Number of Respondents") + 
  theme(axis.text.x = element_text(size = 8, angle = 90))

```



## Your Turn

Using the *forcats* [cheatsheet as a reference](https://posit.co/wp-content/uploads/2022/10/factors-1.pdf), can you: 

  * Modify the code above to collapse the levels of the **marital status** attribute to only include: Married, Never Married, and Other? **Hint:** You can  also use the `fct_collapse` function to accomplish this task. 
  * Can you then generate a similar barchart showing the number of respondents by marital status? 
  * Use the `ggsave` function to export your barchart to a .png file when you are done...

    `ggsave("my_cool_barchart.png", units = "in", width = 16, height = 8)`


```{r label="Your Time To Shine"}


```

<br> 

For the remainder of the session today, review the attributes in the [dataset we have here](https://forcats.tidyverse.org/reference/gss_cat.html) then experiment with some of the commonly used functions from the *dplyr* package that we have seen before (e.g., `filter`, `mutate`, `summarise`). 

When you are done, celebrate `r emo::ji("party")`

<br> 

----


<br> 

<center> You have reached the end! `r emo::ji("beer")` </center>

<br> 

