---
title: "Getting Started with tidyr—Part 2"
date: "Last Updated `r Sys.Date()`"
fontsize: 16pt
output:
  html_document:
    theme: united
    df_print: paged
---
<br> 

This week we are formally introducing the *tidyr* package, which along with *dplyr* provide the backbone and foundation for data wrangling within the *tidyverse*. Let's practice a few of the things you encountered in the assigned reading for today using a familiar dataset: 

 * [Tidy Data](https://r4ds.had.co.nz/tidy-data.html)
 + [Relational Data](https://r4ds.had.co.nz/relational-data.html)

The code chunk below loads libraries we will need into memory and recreates the data object from the preceding Notebook. Because the [chunk option](https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html) `include = FALSE` is there, those lines of code will be run, but **will not** appear in the knitted HTML file. 

```{r label="Load R Packages", include= FALSE}

library(tidyverse)
library(jsonlite)
library(emo)

nyc_ppb <- fromJSON("https://data.cityofnewyork.us/resource/wwhr-5ven.json?%24limit=5000")
nyc_ppb_tb <- as_tibble(nyc_ppb)

```

<br> 

This portion of our work for the day focuses on [pivoting as described in the lead reading](https://r4ds.had.co.nz/tidy-data.html#pivoting). Sometimes we have data with multiple or related characteristics that are **captured in separate rows** or alternately **captured in separate columns** and pivoting allows us to shift between those two ways of structuring the information. Data with a time component are a prime example of when pivoting can come into play. 

In the participatory budgeting tibble, Queens has the most records, so let's extract those and demonstrate how pivoting works by looking at budgeting proposals by category and year. According to the metadata, the borough codes are as follows: 

> Borough Code (1-Manhattan, 2-Bronx, 3-Brooklyn, 4-Queens, 5-Staten Island)

Let's use the `dplyr::select` function to extract the columns that we care about and assign them to a new tibble object. 

<br>

```{r label="Extract Queens Data and Explore Missing Values"}

nyc_ppb_queens <- nyc_ppb_tb %>%
  filter(borough_code == 4) %>%
  select(c(vote_year, category, cost, votes, nta))

dim(nyc_ppb_queens)

nyc_ppb_queens %>% summarise_all(~sum(is.na(.)))


nyc_ppb_queens <- nyc_ppb_queens %>%
  drop_na(nta) 

dim(nyc_ppb_queens)
nyc_ppb_queens %>% summarise_all(~sum(is.na(.)))


```

In the above chunk, we `filter` the records for the borough of Queens, then **only** extract the five columns listed. We check how many rows and columns remain usinbg the base R `dim` function. Next, we display the number of missing data values by column in the resultsing tibble object. There are many ways to do this, but if you want to stay in *dplyr* the code above works. But what is going on with that line of code above? The `summarise_all` function is one of many available in the *dplyr* package for summarizing data and the tilde `~ `specifies a separate function to be applied, which in this case is the combination of `sum` and `is.na`. What follows the tilde is what is called an anonymous function (also known as a lambda expression). This is a function definition that is created and used, but that is never assigned a specific identifier. More on this in two weeks when we cover functions, loops, and iteration.

We have seen how functions can be nested inside one another, but the pipe operator `%>%` is designed to eliminate the need to do that. However, base R functions like `is.na` and `sum` do not work well with the pipe operator (i.e., they are baby boomers to the tidyverse) `r emo::ji("laugh")`. The solution is to use the `.` placeholder to move what is written on the left-hand side (i.e., the budgeting tibble) of the pipe into the first argument of the function on the right-hand side of the pipe (i.e., `summarise_all`). Basically we are saying feed the **nyc_ppb_queens** tibble to the `summarise_all` function as its first argument. Next, we create an ephemeral function that sums the results of the `is.na` function and because the `summarise_all` function operates on all columns in the tibble specified as its first argument, it returns the information we are after. Make sense? `r emo::ji("geek")`

This new tibble is currently in **long format** because we have multiple records with the same year. We can change the orientation of these data using functions from the *tidyr* package `r emo::ji("cool")`... but it is often useful for `summarise` the data first `r emo::ji("wink")`


```{r label="Extract and Pivot Queens Data"}

nyc_ppb_queens_summed <- nyc_ppb_queens %>%
  group_by(category, vote_year) %>%
  mutate(cost_num = as.numeric(cost)) %>%
  summarise(total_cost = sum(cost_num))

dim(nyc_ppb_queens_summed)
glimpse(nyc_ppb_queens_summed)


nyc_ppb_queens_wide <- nyc_ppb_queens_summed %>% 
  pivot_wider(names_from = vote_year, values_from = total_cost)

dim(nyc_ppb_queens_wide)
glimpse(nyc_ppb_queens_wide)


```


### **Your Turn** 

<br>

In the code chunk below, add some code that pivots the built-in [billboard dataset](https://tidyr.tidyverse.org/reference/billboard.html) from **wide format** to **long format** using *tidyr* functions from today's reading. 

<br> 

Just don't ask me why there are **76** weeks rather than **52** weeks... `r emo::ji("eyes")`

<br>

```{r label="Pivot Billboard Hot 100 Data"}

songs_2000 <- billboard

```
<br> 

Oh, [Baha Men](https://youtu.be/ojULkWEUsPs)...

<br> 

...and [Britney Spears](https://youtu.be/CduA0TULnow)...

<br> 

...and [N'Sync](https://youtu.be/Eo-KmOd3i7s)

<br> 

Looks like 2000 was a pretty good music year, no? `r emo::ji("wink")`

<br>

<center> You have reached the end! `r emo::ji("sing")`</center>

<br> 

